---
title: "Data Exploration"
format: html
toc: TRUE
editor_options: 
  chunk_output_type: inline
---


 For the data exploration part, we will try to investigate the cost of living difference in Hawaii, and Illinois by comparing 
 gas price (GASP) and gross rent as percentage of household income (GRPIP) 
```{r}
  # Use the API function to pull data for Illinois, geo_subset=17
explore_il <- single_year(num_vars = c("AGEP", "GRPIP","GASP"),geo_subset = 17)
print(explore_il)

```
```{r}
 # Use the API function to pull data for Hawaii, geo_subset=15 (Default for our group)
explore_hi <- single_year(num_vars = c("AGEP", "GRPIP","GASP"))
print(explore_hi)
```
 Use the summary.census function to calculate the mean gas price and Gross rent as a percentage of household income past 12 months for state Illinois 
 
```{r}
explore_ils <- summary.census(df=explore_il)
explore_ils
class(explore_ils)


```

 Use the summary.census function to calculate the mean gas price and Gross rent as a percentage of household income past 12 months for state Hawaii  
 
```{r}
explore_his <- summary.census(df=explore_hi)
explore_his
class(explore_his)
```
# plot 


```{r}
  # Use the API function to pull data for Illinois, geo_subset=17 and Hawaii geo_subset=15
explore_both <- single_year(num_vars = c("AGEP", "GRPIP","GASP"),geo_subset = c("17","15"))
print(explore_both)

```
```{r}
plot.census(explore_both, "GASP", "ST")
```

```{r}
plot.census(explore_both, "GRPIP", "ST")

```

